---
title: "Churn Analysis" 
output:
  html_document:
    code_folding: "hide"
---

# Business Context

A telecommunication company spent millions by means of services and marketing to attract new customers.  But is there any way for it to make long term relationships with customers? Is there any means to reduce customer churn? More importantly any way to find out what factors influence customer churn? Understanding these factors can help certain company to have an edge over competitors and will ensure growth.

## Scope of analysis:

The conclusion and recommendation of this project should only be applied on customers who satisfy the following criteria: 

* Customers who subscribed to Internet Service. 
* Customers who stayed with us for less than 4 years.

Rationale: Customers who subscribed to Internet Service is the major division in terms of revenue, while customers who stayed with us for less than 4 years have very high churn compared to customers who are with us for more than 4 years.

## Dataset used for analysis

* Description of dataset: Customer data for a telecom company providing home phone and Internet services to 7,043 customers in California during Q3, 2009
* Unit of analysis: a customer staying with the company in the begining of Q3. 
* Source: [Kaggle](https://www.kaggle.com/blastchar/telco-customer-churn)
* Important features:
    + Demographic: Gender, Senior Citizen, Partnered, Dependents
    + Enrollment status of services: Premium Tech Support (**Treatment**),Tenure in Months, Online Backup Plan, Online Security, Streaming TV, PaperlessBilling 
    + Billing: Device Protection Plan
    + Churn Status: Churn Label (**Outcome variable**)
* Other than Tenure in Months, all other attributes are binary variables


### Exploratory Analysis 

**Load data and feature engineering**
```{r echo=TRUE, message=FALSE, warning=FALSE}
library(foreign)
library(ggplot2)
library(dplyr)
library(MatchIt)
churn = read.csv('Telco-Customer-Churn.csv')
#set up scope of analysis
churn_internet = churn %>% filter(InternetService !='No')

#feature engineering
churn_internet$TechSupport = factor(churn_internet$TechSupport,levels = c('No','Yes'))
churn_internet$OnlineSecurity = factor(churn_internet$OnlineSecurity,levels = c('No','Yes'))

churn_internet$OnlineBackup = 
  factor(churn_internet$OnlineBackup,levels = c('No','Yes'))

churn_internet$DeviceProtection = 
  factor(churn_internet$DeviceProtection,levels = c('No','Yes'))

churn_internet$StreamingTV = 
  factor(churn_internet$StreamingTV,levels = c('No','Yes'))

churn_internet$StreamingMovies = 
  factor(churn_internet$StreamingMovies,levels = c('No','Yes'))

churn_internet$DeviceProtection = factor(churn_internet$DeviceProtection,levels = c('No','Yes'))
churn_internet = churn_internet %>% mutate(Contract_new = ifelse(Contract !="Month-to-month",'Year',"Month-to-month"))

churn_internet = churn_internet %>% mutate(is.protected = ifelse(DeviceProtection =="No",0,1))
churn_internet = churn_internet %>% mutate(month_contract =
                                             ifelse(Contract_new == 'Year',0,1))
churn_internet = churn_internet %>% mutate(is.paperless =
                                             ifelse(PaperlessBilling == 'No',0,1))
churn_internet = churn_internet %>% mutate(is.backup =
                                             ifelse(OnlineBackup == 'No',0,1))

churn_internet = churn_internet %>% mutate(is.supported = ifelse(TechSupport == 'No',0,1))

churn_internet_48months = churn_internet %>% filter(tenure>=1 & tenure <=48) 
#Set up treatment
churn_internet_48months$treatment = churn_internet_48months$is.supported  
```


**Descriptive statistics of data**

A. Missingness of data:

* No missing value in the data
```{r}
colSums(is.na(churn_internet_48months))
```

B. Distribution of categorical attributes

```{r}

apply(churn_internet_48months %>% select(!c(customerID,PhoneService,MultipleLines,DeviceProtection,Contract,month_contract,Contract_new,TotalCharges,MonthlyCharges,TotalCharges,Churn,TechSupport,tenure,is.supported,PaymentMethod)) %>% data.matrix,2,table)
```

**Interpretation:** the first row represents 0 (No), the second row represetns 1 (Yes). From the table we can see the distributions of observed attribute's value are not perfectly even except that for gender. 

C. distribution of tenure
```{r}
summary(churn_internet_48months$tenure)
```

**Interpretation:** For customers staying less than 4 years, half of them have a tenure more than 1 year (15 months). On average the tenure is 18 months.  


## Threats to Causal Inference

* Selection Bias - Dataset is provided for Q3 of 2009 and we are making all the inferences based on the subset of a year's data. Before, confirming the causal relationship between variables it is also important to understand if the data provided to us represent the behavior of the all population. It might be the case that customers behave differently during Q3 compared to other quarters. Hence, the causality we establish currently only applied to a period of year and will only be applicable for the quarter we have data for. If we want to generalize the inference, then we need data for the whole year to understand and remove any season-based trends or behavior change in customers.

* Omitted Variable Bias - Presence of omitted variables may result in bias in the GLM model used to establish the causality between customer churn and other variables. In our current data we don't have information about customers demographics like their income, age, qualification levels etc. which might affect their decision to churn from the telecom service provider. Since, we don't have the information about these features; hence, there is a threat that model might result in bias. The most important assumption we are making here is that people in urban and rural areas behave similarly as we don't have data that demarcates user into these groups. But there is high threat of varying user behavior between urban and rural areas.

* Reverse Causality - It is really important to understand that if customer churning is caused by not providing technical support or a decrease in support is observed because customers are churning. It is important to understand how the causality is related as methods used in our analysis does not specifically specifies the direction of causality. Hence, the reverse direction of causality although weak but can be considered a threat. 

* Threat of Matching  Process - Currently matching is performed only using only the treatment variable which is whether the technical support is provided to user or not. Hence, while drawing inferences about causal affect of other variables like whether the customer is enrolled in paperless billing and streaming TV will have to made with caution. If we want to increase the confidence of such claims then a separate matching based on these variable should be performed before using data in the GLM model. Also, the use of logit function for matching is somewhat arbitrarily; hence, to further confirm if the logit function is actually resulting in right approximation of observed variables; it should be checked with other advanced machine learning methods. 

## Method

### Randomization Test:

Since this is an observational data, there's no guanrantee that the treatment group and test group are similar except the value of treatment. Therefore a randomization test is conducted for each observed attribute. 

Moreover, since most of the attributes are categorical, A **chi-square test** is performed in this case to see whether the distribution of an observed attribute is different among treatment groups. For the continuous attribute `tenure in month`, a regular t-test is performed.

Besides randomization Test, a comparison among distributions of propensity score for different tratment groups is also performed to reflect the validity of randomization assumption.

**Code and results of randomization tests**

```{r echo=TRUE}
churn_internet_beforematch = churn_internet_48months %>% select(gender,SeniorCitizen,
                                            Partner,
                                            Dependents,
                                            is.paperless,
                                            is.protected,
                                            is.backup,
                                            #TechSupport,
                                            OnlineSecurity,
                                            StreamingTV,
                                            StreamingMovies,
                                            tenure,
                                            treatment,
                                            Churn)
#chi-square test for categorical variable
p_value_b = function(var){
  return(chisq.test(var,churn_internet_beforematch$treatment)$p.value)
}
apply(churn_internet_beforematch %>% select(!c(Churn,treatment,tenure)),2,p_value_b)

# t-test
t.test(churn_internet_beforematch$tenure~churn_internet_beforematch$treatment)
```


**Compare the distribution of propensity scores**
```{r echo=TRUE}


ps_before = glm(treatment ~as.factor(gender)+
                          as.factor(SeniorCitizen)+
                          as.factor(Partner)+
                          as.factor(Dependents)+
                          as.factor(is.paperless)+
                          as.factor(is.backup)+
                      #    as.factor(TechSupport)+
                          as.factor(is.protected)+
                  as.factor(OnlineSecurity)+
                  as.factor(StreamingMovies)+
                  as.factor(StreamingTV)+
                          tenure, data = churn_internet_beforematch,
    family = binomial)

churn_internet_beforematch$PS = ps_before$fitted.values

ggplot(churn_internet_beforematch,aes(x=PS,color = factor(treatment)))+geom_density()+ggtitle('Distributions of propensity scores before matching')+scale_color_discrete(name = "Treatment", labels = c("No Tech Support","Has Tech Support"))+xlab('Propensity Score')+theme_classic()
```

**Analysis of randomization test:** most randomization test return significant p-value, suggesting that the distributions of certain observed attribute are not similar among treatment groups. 

**Conclusion of randomization test:** the data is not randomized for making causal inference. 

### Matching: Propensity Score Approach

Since the assumption of randomization fails, we use matching technique (propensity score) to identify pairs which are similar on all characteristics except treatment - premium tech support and define them as treatment and control groups.

**Assumption of propensity score approach:** the treatment assignment is fully determined by observed attributes.

**parameters of matching: **
* method = nearest
* calibre = 0.01
* replace = FALSE
* ratio = 1

**Code of matching**
```{r}
match_output <- matchit(treatment ~as.factor(gender)+
                          as.factor(SeniorCitizen)+
                          as.factor(Partner)+
                          as.factor(Dependents)+
                          as.factor(is.paperless)+
                          as.factor(is.backup)+
                      #    as.factor(TechSupport)+
                          as.factor(is.protected)+
                  as.factor(OnlineSecurity)+
                  as.factor(StreamingMovies)+
                  as.factor(StreamingTV)+
                          tenure
                          , data = churn_internet_beforematch, 
                        method = "nearest", distance = "logit", calibre=0.01,
                        replace = FALSE)

churn_internet_aftermatch = match.data(match_output)

ps_after = glm(treatment ~as.factor(gender)+
                          as.factor(SeniorCitizen)+
                          as.factor(Partner)+
                          as.factor(Dependents)+
                          as.factor(is.paperless)+
                          as.factor(is.backup)+
                      #    as.factor(TechSupport)+
                          as.factor(is.protected)+
                  as.factor(OnlineSecurity)+
                  as.factor(StreamingMovies)+
                  as.factor(StreamingTV)+
                          tenure, data = churn_internet_aftermatch,
    family = binomial)

churn_internet_aftermatch$PS = ps_after$fitted.values
```

**randomization test and a comparison of propensity score distributions after matching**
```{r}
#chi square test
p_value = function(var){
  return(chisq.test(var,churn_internet_aftermatch$treatment)$p.value)
}
apply(churn_internet_aftermatch %>% select(!c(Churn,PS,distance,weights,treatment,tenure)),2,p_value)

#t test
t.test(churn_internet_aftermatch$tenure~churn_internet_aftermatch$treatment)
```

**compare the propensity score distribution**
```{r echo=FALSE}

ggplot(churn_internet_aftermatch,aes(x=PS,color = factor(treatment)))+geom_density()+ggtitle('Distributions of propensity scores after matching')+scale_color_discrete(name = "Treatment", labels = c("No Tech Support","Has Tech Support"))+xlab('Propensity Score')+theme_classic()
```

**Change in sample size after matching**
```{r echo=TRUE}
paste("# of sample in each treatment group before matching")
(table(churn_internet_beforematch$treatment))

paste("# of sample in each treatment group after matching")
(table(churn_internet_aftermatch$treatment))
```

**Interpretation of matching result:** All randomization test returns insignificant p-values (as large as 0.11). The distributions of propensity scores after matching are also mostly overlapped based on visualization above. The matching is successful in terms of the fullfillment of randomization assumption.

### Sensitivity Analysis of Matching: 

The key assumption of matching is that there is no omitted bias in the data. To validate this assumption, a sensitivity analysis of matching is conducted after major modeling. **(See the second part of Modeling section)**

## Modeling

### main modeling

For the modeling part we are using all the variables. Since matching has worked the current data is as close as possible to an experiment and hence using all the variables in the model would ensure we are capturing the true effect of the treament with minimum standard deviation. In addition, many confounding variables are also levers that can be pulled and therefore it is reasonable to take a look at their potential effects on the churn out decision.

**Result of modeling**
```{r echo=TRUE}
log = glm(Churn ~ as.factor(treatment)+as.factor(gender)+
                          as.factor(SeniorCitizen)+
                          as.factor(Partner)+
                          as.factor(is.backup)+
                          as.factor(Dependents)+
                          as.factor(is.paperless)+
                          #as.factor(TechSupport)+
                          as.factor(is.protected)+
                          as.factor(OnlineSecurity)+
                          as.factor(StreamingMovies)+
                          as.factor(StreamingTV)+
                          tenure, data = churn_internet_aftermatch,
    family = binomial)

summary(log)
```
**Interpretation and conclusion of modeling result:** 

* From the above model we are seeing that if a customer starts using the preimum tech support feature, there is an decrease in logodds by 0.64 and hence the probability of churning is also decreased. 

* Also we noticed a high positive coefficient value on the paperless billing / Streaming TV usage which translates to customers who opted for paperless billing or use company's service to watch streaming TV may be churning more but we cannot say this with full confidence as the data is matched based on premium tech support.


### Sensitivity Analysis of Matching

After multiple trials, the following alternative setting of parameters returns a valid matching: 

* distance = 'probit' (compared with "logit" in original setting)
* method = nearest
* calibre = 0.01
* replace = FALSE
* ratio = 1

**Result of matching:**
```{r echo=TRUE}
match_output_sen2 <- matchit(treatment ~as.factor(gender)+
                          as.factor(SeniorCitizen)+
                          as.factor(Partner)+
                          as.factor(Dependents)+
                          as.factor(is.paperless)+
                          as.factor(is.backup)+
                      #    as.factor(TechSupport)+
                          as.factor(is.protected)+
                  as.factor(OnlineSecurity)+
                  as.factor(StreamingMovies)+
                  as.factor(StreamingTV)+
                          tenure
                          , data = churn_internet_beforematch, 
                        method = "nearest", distance = "probit", calibre=0.01,
                        replace = FALSE)

churn_internet_aftermatch_sen2 = match.data(match_output_sen2)

ps_after_sen2 = glm(treatment ~as.factor(gender)+
                          as.factor(SeniorCitizen)+
                          as.factor(Partner)+
                          as.factor(Dependents)+
                          as.factor(is.paperless)+
                          as.factor(is.backup)+
                      #    as.factor(TechSupport)+
                          as.factor(is.protected)+
                  as.factor(OnlineSecurity)+
                  as.factor(StreamingMovies)+
                  as.factor(StreamingTV)+
                          tenure, data = churn_internet_aftermatch_sen2,
    family = binomial)

churn_internet_aftermatch_sen2$PS = ps_after_sen2$fitted.values

ggplot(churn_internet_aftermatch_sen2,aes(x=PS,color = factor(treatment)))+geom_density()+ggtitle('Distributions of propensity scores after matching (replace = True)')+scale_color_discrete(name = "Treatment", labels = c("No Tech Support","Has Tech Support"))+xlab('Propensity Score')+theme_classic()
```

**Randomization test for each observed attribute**
```{r}
p_value2 = function(var){
  return(chisq.test(var,churn_internet_aftermatch_sen2$treatment)$p.value)
}
apply(churn_internet_aftermatch_sen2 %>% select(!c(Churn,PS,distance,weights,treatment,tenure)),2,p_value2)


t.test(churn_internet_aftermatch_sen2$tenure~churn_internet_aftermatch_sen2$treatment)
```

**Modeling under alternative matching result**
```{r}
log_a = glm(Churn ~ as.factor(treatment)+as.factor(gender)+
                          as.factor(SeniorCitizen)+
                          as.factor(Partner)+
                          as.factor(is.backup)+
                          as.factor(Dependents)+
                          as.factor(is.paperless)+
                          #as.factor(TechSupport)+
                          as.factor(is.protected)+
                          as.factor(OnlineSecurity)+
                          as.factor(StreamingMovies)+
                          as.factor(StreamingTV)+
                          tenure, data = churn_internet_aftermatch_sen2,
    family = binomial)

summary(log_a)
```

**Analysis of sensitivity test result: ** With an alternative setting of matching parameters, the model detects the treatment effect on the variable, and the estimate of the scale of effect is comparable. (-0.62 vs -0.6)

**Conclusion of sensitivity analysis: ** The interpretation and conclusion of the modeling result does not change with another setting of matching parameters, indicating the robustness of our analysis.

##Recommendation:

Based on the first conclusion of main modeling section, we recommend company to start promoting the premium tech support more so that more customers enroll in this service and reduce the chance of churn.(Company should conduct a sound cost benefit analysis in advance to make sure that the ROIs of these solutions are positive.)

* Advertise more about the premium tech support
* offer premium tech support for a discount to encourage more customers to join.
  
Based on the second conclusion of main modeling section, we recommend company to do more experiments to test the true impact of paperless billing / Stremaing TV usage on churn respectively.





